%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% aiih_nova_prompting.tex   (LLNCS, natbib numeric)  -  tool-use focus, tidy style
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[runningheads]{llncs}

% ----------------------------------------------------------------------------- 
% Essential packages
% -----------------------------------------------------------------------------
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage[numbers]{natbib}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,positioning,fit,backgrounds}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref}   % keep last

% ----------------------------------------------------------------------------- 
% Title and authors
% -----------------------------------------------------------------------------
\title{Agentic Reasoning in Vision-Language Models\\
       for Rare Brain MRI Analysis: A Multi-Turn Tool-Augmented Approach}

\titlerunning{Agentic VLM Reasoning on NOVA}

\author{%
Duaa Alim\inst{1}\thanks{Corresponding author: \email{d.alim@imperial.ac.uk}}%
\and Lina Alim\inst{2}%
\and Liam Chalcroft\inst{3}}

\authorrunning{Alim \textit{et al.}}

\institute{Imperial College London, UK
\and Hillingdon Hospitals NHS Foundation Trust, UK
\and University College London, UK}

\begin{document}
\maketitle

% ----------------------------------------------------------------------------- 
% Abstract
% -----------------------------------------------------------------------------
\begin{abstract}
Vision-language models struggle with rare brain pathologies that require specialist medical knowledge insufficiently learned during training.  
We investigate agentic reasoning enabling autonomous analysis through iterative tool selection and evidence integration.  
Five configurations are compared on the NOVA benchmark~\cite{bercea2025novabenchmarkanomalylocalization}: (i) baseline single-shot analysis, (ii) autonomous multi-turn reasoning, (iii) web search-augmented reasoning, (iv) visual operations-guided analysis, and (v) comprehensive tool-augmented agents.  
Our framework enables autonomous decisions about reasoning continuation, tool usage (medical web search, visual operations), and evidence synthesis across iterative turns.  
Results show agentic reasoning significantly outperforms baseline approaches, with the comprehensive agent achieving 90\% improvement in localisation mAP$_{50}$ (13.7 vs 7.2) while maintaining competitive caption quality.
\end{abstract}

\keywords{Vision-language models \and Medical imaging \and Agentic reasoning \and Tool-augmented agents \and Rare pathology}

% =============================================================================
\section{Introduction}
% =============================================================================
Current vision-language models fail catastrophically on rare brain pathologies requiring specialist medical knowledge these models insufficiently acquire during training. The NOVA benchmark~\cite{bercea2025novabenchmarkanomalylocalization} reveals this failure: 900 brain MRI scans spanning 281 rare neurological conditions expose dramatic performance degradation in state-of-the-art models. This matters because rare diseases collectively affect 400 million people globally, yet the specialized expertise needed for their analysis receives minimal attention in AI development.

We address this critical gap through agentic reasoning, enabling models to autonomously direct their analysis via iterative tool selection and evidence integration. Unlike single-shot inference, our agents make deliberate decisions about when to search medical literature, apply visual operations, or continue reasoning, transforming models from passive pattern matching to active investigation that can acquire and apply specialist knowledge.

Our work evaluates caption generation and abnormality localisation, setting aside diagnosis evaluation due to its complex semantic matching requirements.

% =============================================================================
\section{Related work}
% =============================================================================
\subsection{Medical VLM benchmarks and existing approaches}
Medical VLM evaluation has evolved beyond simple accuracy metrics. CARES~\cite{xia2024carescomprehensivebenchmarktrustworthiness} measures hallucination, bias and calibration; MultiMedEval~\cite{royer2024multimedevalbenchmarktoolkitevaluating} aggregates multi-task performance. However, existing adaptations like LLaVA-Med~\cite{li2023llavamedtraininglargelanguageandvision} and XrayGPT~\cite{thawakar2025xraygptchestradiographssummarization} focus on common clinical images, leaving rare disorders largely unaddressed.

\subsection{Agentic reasoning and autonomous decision-making}
External tools unlock emergent reasoning capabilities in large language models~\cite{berti2025emergentabilitieslargelanguage}. Chain-of-thought and visual prompting demonstrate reduced hallucination in medical VQA~\cite{Guo_2025,zhu2025guidingmedicalvisionlanguagemodels}. Recent surveys highlight potential for autonomous analytical processes~\cite{wang2024promptengineeringhealthcaremethodologies, chen2024surveymedicalvisionandlanguageapplications}, yet few studies explore autonomous tool selection in medical imaging. Agentic frameworks enable autonomous determination of reasoning continuation, tool usage, and evidence synthesis strategies with clinical context integration.

% =============================================================================
\section{Methodology}
% =============================================================================
\subsection{Dataset}
The NOVA benchmark~\cite{bercea2025novabenchmarkanomalylocalization} provides brain MRI scans across multiple sequences (T1, T2, FLAIR) with expert annotations for rare neurological pathologies. We evaluate on the public test split without model fine-tuning, focusing on out-of-distribution generalisation.

\subsection{Agentic reasoning configurations}
We systematically compare five configurations on the NOVA benchmark, each building upon the previous:
\begin{enumerate}
  \item \textbf{Baseline}: Single-shot analysis integrating clinical context with structured output generation.
  \item \textbf{Autonomous Reasoning}: Multi-turn reasoning where agents decide when to continue analysis and how to synthesise evidence.
  \item \textbf{Web-Augmented Agent}: Autonomous reasoning enhanced with medical web search capabilities, allowing agents to retrieve guidelines, research, and anatomical information as needed.
  \item \textbf{Visual Operations Agent}: Autonomous reasoning with visual tools for spatial manipulations (zoom, crop, contrast adjustment) and coordinate validation.
  \item \textbf{Comprehensive Agent}: Full autonomy integrating all capabilities, with agents selecting optimal tools and reasoning strategies across multiple turns.
\end{enumerate}

\subsection{Tasks and evaluation metrics}
We evaluate two core tasks: \textbf{Caption} generation of comprehensive radiological descriptions, evaluated using METEOR which measures semantic similarity through alignment of stems, synonyms, and paraphrases; and \textbf{Localisation} of abnormalities with precise bounding boxes, evaluated using mean average precision (mAP) at multiple IoU thresholds: mAP$_{30}$ (IoU $\geq$ 0.3), mAP$_{50}$ (IoU $\geq$ 0.5), and mAP$_{50\text{-}95}$ (averaged across IoU thresholds 0.5 to 0.95 in 0.05 increments). IoU measures spatial overlap: $\text{IoU} = \frac{\text{Area of Intersection}}{\text{Area of Union}}$.

\subsection{Implementation details}
Our agentic reasoning framework is implemented as a modular Python system with Hydra configuration management. Model access is provided through the OpenRouter API, with comprehensive evaluation conducted using Google Gemini 2.5 Flash~\cite{geminiteam2025geminifamilyhighlycapable}. Prompt templates are implemented using Jinja templating engine, enabling flexible and maintainable prompt engineering across different reasoning configurations.

The framework enables autonomous decision-making through structured reasoning schemas, clinical context integration, spatial reasoning with coordinate validation, and evidence integration across turns. Tool capabilities include curated medical web search and coordinate-validated visual operations, with systematic logging and automated evaluation metrics computation.

% =============================================================================
\section{Results and Discussion}\label{sec:results}
% =============================================================================

\begin{figure}[t]
  \centering
  \begin{tikzpicture}[
    node distance = 1cm,
    % -----------------------------------------------------------------
    reasoning/.style={rectangle, draw, fill=blue!20,
                      text width=1.6cm, align=center, minimum height=1cm},
    decision/.style ={diamond,   draw, fill=orange!20,
                      text width=1.4cm, align=center,
                      shape aspect=2,   % squish vertically
                      inner sep=1pt},
    tool/.style     ={rectangle, draw, fill=green!20,
                      text width=1.2cm, align=center, minimum height=0.8cm},
    input/.style    ={ellipse,   draw, fill=gray!20,
                      align=center, minimum width=1.4cm, minimum height=0.8cm},
    output/.style   ={ellipse,   draw, fill=purple!20,
                      align=center, minimum width=1.4cm, minimum height=0.8cm},
    arrow/.style    ={->, >=stealth, thick},
    dasharrow/.style={->, >=stealth, thick, dashed},
    plate/.style    ={rectangle, draw, dashed, inner sep=0.4cm}
  ]
  % -------------------------------------------------------------------
  % Main nodes
  \node[input] (input) {Brain MRI};

  \node[reasoning, right=1.1cm of input] (reason) {Reasoning\\step};

  \node[decision,  right=1.1cm of reason] (decide) {Continue?};

  \node[output,    right=1.8cm of decide] (tasks) {Task\\execution};

  % Tool node above the diamond
  \node[tool, above=0.6cm of decide] (tool) {Tool};

  % Plate and iteration note
  \node[plate, fit=(reason) (decide) (tool)] (plate) {};
  \node[below=0.2cm of plate.south, font=\small\itshape] {$N \le 3$ iterations};

  % -------------------------------------------------------------------
  % Flow arrows
  \draw[arrow] (input)  -- (reason);
  \draw[arrow] (reason) -- (decide);
  \draw[arrow] (decide) -- node[above,font=\small]{Finish} (tasks);

  % Tool arrow
  \draw[dasharrow] (decide) -- (tool);
  \end{tikzpicture}
  \caption{Agentic reasoning workflow.  The agent may perform up to
           \(N \le 3\) reasoning steps, deciding each time whether to
           continue or finish.  At any decision point it can invoke
           external tools (web search or visual operations).}
  \label{fig:framework}
\end{figure}


\begin{table}[t]
  \caption{Mean ($N=906$) performance comparison on NOVA benchmark (higher is better for all metrics).}
  \label{tab:performance}
  \centering
  \begin{tabular}{lcccc}
    \toprule
    \textbf{Agent Configuration} & \textbf{mAP$_{30}$} & \textbf{mAP$_{50}$} & \textbf{mAP$_{50\text{-}95}$} & \textbf{METEOR}\\
    \midrule
    Baseline                   & 21.6 & 7.2 & 2.5 & 19.0\\
    Reasoning                  & 28.5 & 10.1 & 3.7 & 18.3\\
    Reasoning + Web            & 28.4 & 11.4 & 3.9 & 19.3\\
    Reasoning + Visual         & 28.8 & 9.8 & 4.0 & 17.7\\
    Reasoning + Web + Visual   & \textbf{31.0} & \textbf{13.7} & \textbf{4.6} & \textbf{19.8}\\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{Agentic reasoning effectiveness.}
Our comprehensive agent delivers substantial performance gains across all localisation metrics: mAP$_{30}$ of 31.0, mAP$_{50}$ of 13.7, and mAP$_{50\text{-}95}$ of 4.6, representing 43\%, 90\%, and 84\% improvements over baseline (21.6, 7.2, and 2.5 respectively). For caption generation, the comprehensive agent achieves the best METEOR score of 19.8, with the web-augmented agent also showing strong performance at 19.3, both outperforming baseline's 19.0.

Statistical analysis confirms key improvements are significant. Paired $t$-tests with false discovery rate correction ($\alpha = .05$) reveal: the comprehensive agent significantly outperforms baseline in localisation (mAP$_{30}$: $p = .0008$, mAP$_{50}$: $p = .0017$); and baseline significantly outperforms visual agent in caption generation (METEOR: $p = .0046$).

% =============================================================================
\section{Conclusion and future work}
% =============================================================================
This work demonstrates that agentic reasoning transforms vision-language model performance on rare brain pathologies without requiring model fine-tuning. Our autonomous agents achieve breakthrough performance by directing their own analytical processes: iteratively selecting tools, integrating evidence, and making strategic decisions about when to continue reasoning. The comprehensive agent's 90\% improvement in precise localisation (mAP$_{50}$) represents a significant advance for rare pathology analysis.

These findings have broader implications for medical AI: autonomous analytical processes may prove essential for handling the long tail of rare conditions that traditional training approaches cannot address. Future work will explore diagnosis evaluation, evaluation on reasoning post-trained models~\cite{kumar2025llmposttrainingdeepdive}, sophisticated agentic architectures, and extension across diverse medical imaging domains.

% =============================================================================
\begin{credits}
\subsubsection*{Acknowledgements}
We thank Cosmin Bercea \textit{et al.} for the NOVA benchmark and the medical AI community for establishing standardised evaluation frameworks for rare pathology analysis.

\subsubsection*{Disclosure of interests}
The authors declare no competing interests.
\end{credits}

% -----------------------------------------------------------------------------
\bibliographystyle{splncs04}
\bibliography{references}
\end{document}